Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-30 report/MSc_Report.tex
.

% Checklist: integrate unified transport framing, enforce US spelling, add dynamic refs.
Multivariate density estimation supports likelihood-based modeling, anomaly detection, simulation, and decision making under uncertainty. Tabular datasets often combine moderate to high dimension with context-dependent conditional structure. Conditional variance can change with predictors, and conditional skewness or modality can depend on earlier variables. A transport perspective addresses these challenges by coupling the target to a simple reference through an invertible map. This perspective enables exact likelihoods, transparent conditionals, and efficient sampling through a shared computational backbone \citep{rosenblatt1952remarks,knothe1957contributions}. This thesis compares three estimator families inside a single evaluation frame with matched objectives and diagnostics. The frame evaluates lower-triangular transport maps, Transformation Random Forests interpreted via probability integral transforms, and copulas that decouple marginals from dependence \citep{hothorn2017transformation,hothorn2018conditional,sklar1959fonctions}.
\medskip

We adopt a shared evaluation protocol summarized in Section~\ref{sec:ch1-frame} and detailed in Chapter~\ref{ch:background}; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the pipeline at a glance.
\medskip

For clarity, efficiency, and interpretability we use separable triangular components as in Equation~\eqref{eq:transport-separable}. This structure stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.

% Checklist: add thesis overview, enforce dynamic references, keep ASCII hyphenation.
\section{Thesis and Problem Statement}\label{sec:ch1-problem}
This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame. We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copula baselines.
\medskip

A transport perspective couples standardized data to a Gaussian reference through a monotone lower-triangular map. This structure yields exact likelihoods, transparent conditionals, exact inversion by back substitution, and linear per-sample evaluation. The Rosenblatt and Knothe rearrangements justify the triangular coupling for any variable order \citep{rosenblatt1952remarks,knothe1957contributions}.
\medskip

We standardize features using training statistics only. Equation~\eqref{eq:transport-standardise} defines the standardized coordinates $u$ used for evaluation. All derivatives and Jacobians are computed in $u$. The diagonal affine correction in Equation~\eqref{eq:transport-affine} reports log densities on the original scale $x$. This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets. All log quantities are reported in nats.
\medskip

We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\varphi$ and $\Phi$. Abbreviations for models and references appear in Table~\ref{tab:model-abbrev}. Copulas decouple marginals from dependence and serve as interpretable baselines.
\medskip

Separable triangular maps decompose each component into a context shift and a univariate monotone shape as in Equation~\eqref{eq:transport-separable}. The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant. Under strictly increasing conditional CDFs after standard monotone smoothing and with an additive predictor, TRTF implements the same separable triangular likelihood via the probability integral transform. Copulas preserve explicit marginals and model dependence on the unit hypercube; in this thesis they serve strictly as low-dimensional ($K\!\le\!3$) diagnostic baselines and are not evaluated on high-$K$ datasets.

We evaluate all estimators under the single protocol referenced above, with matched preprocessing and reporting to keep results comparable. Section~\ref{sec:evaluation-protocol} defines metrics and timing conventions.

Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes the pipeline by showing standardization $u = T_{\text{std}}(x)$, the triangular transport branch containing TTM-Sep and TRTF, and the copula branch. Both branches feed the reported outputs, namely log density, conditionals, sampling, calibration, and compute, under the shared frame.

The central problem is to determine when separability is appropriate for tabular data. We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention (reported log densities on $x$ apply the affine correction in Eq.~\eqref{eq:transport-affine}). Ordering effects, conditional calibration, and computational trade-offs address this question.
\medskip

On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MiniBooNE benchmark it improves on Gaussian references but trails published flow baselines. Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.

\paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions. We also exclude non-separable TRTF predictors and cross-term triangular maps because they raise compute and tuning costs substantially and complicate calibration/identifiability in our setting; the present scope focuses on separable structure for transparency and exact inversion. Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference. A brief pointer to expected changes if these variants were enabled—how cross-terms or non-additive predictors would likely affect calibration and NLL—is given in Chapter~\ref{ch:conclusion}, Section~\ref{sec:conclusion-outlook}.

% Checklist: consolidate notation, relocate figure, keep one-claim sentences.
\section{The Transport Frame on One Page}\label{sec:ch1-frame}
To avoid duplication, the canonical derivations and notation live in Chapter~\ref{ch:background}. This section serves only as a map: we standardize with train-only statistics (Eq.~\eqref{eq:transport-standardise}), evaluate likelihoods via the pullback (Eq.~\eqref{eq:transport-pullback}), exploit the triangular determinant factorization (Eq.~\eqref{eq:transport-det}), and apply the affine correction for reporting (Eq.~\eqref{eq:transport-affine}). Separable components are defined in Eq.~\eqref{eq:transport-separable}. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the pipeline.

Notation remains consistent. We write $\eta$ for the $K$-variate standard normal density, and $\varphi$ and $\Phi$ for the univariate standard normal density and CDF. We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$. These choices align symbols across Chapters~\ref{ch:intro}--\ref{ch:dataanalysis} and prevent ambiguity in later diagnostics and tables.

This one-page frame removes duplicated exposition from Chapter~\ref{ch:background}. It establishes where logs and Jacobians live and makes complexity, inversion, and units explicit before the comparisons that follow. Section~\ref{sec:ch1-problem} documented the motivation, and Section~\ref{sec:ch1-contributions} states the resulting contributions and research questions.

% Checklist: map contributions to evidence, keep bridges to later sections, ensure dynamic refs.
\section{Contributions and Research Questions}\label{sec:ch1-contributions}
This section states the contributions and the research questions, and maps them to the chapters and figures that deliver the evidence. We adopt the shared transport frame summarized in Section~\ref{sec:ch1-frame}; Chapter~\ref{ch:background} records notation and assumptions, and Figure~\ref{fig:transport-schematic} anchors the comparisons.

The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests, and copula baselines. Where we claim an equivalence between TRTF and separable triangular transports, it holds under the conditions made explicit in Section~\ref{sec:transport-trtf} (strictly increasing conditional CDFs after monotone smoothing and an additive predictor). Chapter~\ref{ch:background} and Figure~\ref{fig:transport-schematic} establish the conventions and remove duplication in later chapters.

The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting. We evaluate TTM-Sep, TTM-Marg, and TRTF on synthetic generators and real tabular data; copulas are included only as low-dimensional ($K\!\le\!3$) diagnostic baselines (e.g., Half-Moon, 4D) and are not used in high-$K$ studies. The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation. Section~\ref{sec:evaluation-protocol} defines the protocol, Section~\ref{sec:synthetic-results} presents the synthetic and autoregressive results, and Section~\ref{sec:realdata} positions our measurements against published normalizing-flow baselines where appropriate.

The third contribution distills practical guidance from the unified frame and the benchmarks. We state operational choices that preserve comparability, highlight ordering sensitivity and separability limits, and summarize when copulas serve as informative baselines. Chapter~\ref{ch:conclusion} consolidates these points as actionable recommendations and records limitations that motivate richer parameterizations or alternative predictors.

Two questions drive the empirical study and bind the contributions to specific measurements. The first question asks how TRTF compares with TTM-Sep and copula baselines on synthetic data. All estimators share the transport frame in this comparison. We answer by reporting average test negative log-likelihoods, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost. Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.

The second question asks how closely our TRTF results on real benchmarks approach the published performance of modern normalizing flows under the standard preprocessing. We answer by placing our test log-likelihoods beside reported numbers from the literature. The gaps are interpreted through the separable Jacobian constraint and compute profiles. Section~\ref{sec:realdata} reports these comparisons, and Chapter~\ref{ch:conclusion} interprets their implications for model choice.

Scope and non-goals maintain focus and ensure reproducibility. We study separable triangular maps and TRTF with additive predictors and compute all derivatives and Jacobians in standardized coordinates. The map direction $S:u \to z$ remains fixed for evaluation and inversion. Copulas include Gaussian dependence and a low-dimensional nonparametric variant used strictly as a diagnostic baseline. We treat high-capacity flows as external references rather than primary models, and we do not evaluate non-additive TRTF variants or cross-term triangular maps in this chapter. Chapter~\ref{ch:background} records the formal assumptions and notation. Section~\ref{sec:datasets-preprocessing} details preprocessing, seeds, and reporting conventions that keep results comparable across datasets and estimators.

Taken together, these commitments make the comparisons interpretable, keep units and complexity explicit, and prepare the reader for the empirical evidence that answers the two questions under a single, transparent evaluation frame.

% Abbreviation table for reuse across chapters
\begin{table}[t]
  \centering
  \caption{Model abbreviations used throughout the thesis.}
  \label{tab:model-abbrev}
  \begin{tabular}{ll}
    \hline
    Label & Meaning \\
    \hline
    TTM-Marg & Marginal triangular transport map (per-dimension; no context) \\
    TTM-Sep & Separable triangular transport map (additive: $g_k$ shift + monotone $h_k$) \\
    TRTF & Transformation Random Forests (axis-parallel splits) \\
    True-Marg & Oracle marginal density \\
    True-Joint & Oracle conditional joint density \\
    Copula & Copula baseline (Gaussian or nonparametric) \\
    \hline
  \end{tabular}
\end{table}

@@

\section{Transport Frame and Notation}\label{sec:transport-frame}

This section fixes the standardized coordinate system, notation, and algebraic identities used throughout the thesis. The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters. We summarize the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorization that drives evaluation and inversion.

We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
\begin{equation}
  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,\label{eq:transport-standardise}
\end{equation}
where $\mu$ and $\sigma$ denote the empirical mean and standard deviation estimated on the training split and $\oslash$ denotes elementwise division. In words, we shift and rescale features once, using training data only, and keep all derivatives and Jacobians in $u$-space to avoid leakage and to ensure comparability across estimators.

The standardized density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$. Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$. The pullback identity then reads
\begin{equation}
  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|,\label{eq:transport-pullback}
\end{equation}
which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant. Reporting log densities on the original scale requires only the diagonal affine correction implied by standardization,
\begin{equation}
  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.\label{eq:transport-affine}
\end{equation}
We therefore differentiate with respect to $u$, and we convert to $x$-scale only at reporting time.

The transport is assumed to be lower triangular and componentwise monotone,
\begin{equation}
  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), \qquad \partial_{u_k}S_k(u_{1:k})>0,\label{eq:transport-triangular}
\end{equation}
so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorizes into a sum of one-dimensional log derivatives,
\begin{equation}
  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).\label{eq:transport-det}
\end{equation}
The factorization yields $\mathcal{O}(K)$ evaluation cost per-sample, improves numerical stability, and guarantees global invertibility: strictly monotone diagonal derivatives let us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.

Table~\ref{tab:transport-notation} consolidates the notation used in this transport frame. All derivatives and Jacobians act on $u$; the affine correction \eqref{eq:transport-affine} converts log densities back to $x$ for reporting. The remainder of this chapter adopts this frame. Section~\ref{sec:transport-separable} details the separable triangular parameterization used for direct transports. Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform under the conditions stated there (strictly increasing conditional CDFs after monotone smoothing and an additive predictor). Section~\ref{sec:transport-copula} places copulas in the same reporting convention.

\begin{table}[t]
  \centering
  \caption{Notation for the transport frame used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}. All derivatives and Jacobians are taken with respect to $u$; log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.}
  \label{tab:transport-notation}
  \begin{tabular}{ll}
    \hline
    Symbol & Meaning \\ 
    \hline
    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ \\ 
    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates \\ 
    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
    $\eta(z)$ & $K$-variate standard normal density \\ 
    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space, respectively \\ 
    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
    $K$ & Dimension of the feature vector \\ 
    \hline
  \end{tabular}
\end{table}

\section{Separable Triangular Maps and Transformation Random Forests as Transport}\label{sec:transport-separable}

% Checklist: enforce US spelling, ensure dynamic cross-references, keep equation labels aligned.
This section unifies separable triangular maps and Transformation Random Forests (TRTF) within the transport frame fixed in Section~\ref{sec:transport-frame}. Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$. The use of triangular transports builds on modern measure-transport literature; see, for instance, triangular transformations and their properties in \citet{bogachev2005triangular}. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}. We focus on shared likelihood identities, modeling assumptions, and limits of separability, and defer implementation details to Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.

The goal is to state a single likelihood for both constructions, clarify what separability permits, and identify failure modes that motivate richer parameterizations. We do not pursue non-additive TRTF predictors, cross-term triangular maps, or ordering heuristics in this section; Chapter~\ref{ch:dataanalysis} evaluates those choices empirically and Appendix~\ref{ch:appendix} documents routines and defaults.

We adopt the notation introduced in Section~\ref{sec:transport-frame}. Coordinates satisfy $u=T_{\mathrm{std}}(x)$, the reference density is $\eta(z)$, and the pullback identity \eqref{eq:transport-pullback} gives $\pi_U(u)=\eta(S(u))\,|\det\nabla_u S(u)|$. The map is lower-triangular with strictly positive diagonal partial derivatives, which yields the sum decomposition in Equation~\eqref{eq:transport-det}. These conventions keep derivatives in $u$-space and apply the affine correction \eqref{eq:transport-affine} only when reporting $\log \pi_X(x)$.

We restrict attention to separable triangular maps. Component $k$ decomposes into a context shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k),\qquad \log \partial_{u_k}S_k(u_{1:k}) \;=\; \log h_k'(u_k),\label{eq:transport-separable}
\end{equation}
which fixes context effects in $g_k$ and reserves $h_k$ for the one-dimensional marginal shape. Intuitively, earlier coordinates translate the location, while the conditional shape along $u_k$ remains fixed across contexts. The Jacobian contribution depends only on $u_k$, which reduces per-sample evaluation cost and simplifies inversion.

\begin{shaded}
\textbf{Assumptions.} Unless stated otherwise, we assume:
\begin{itemize}
  \item \emph{Lower-triangularity:} $S$ has the structure in Eq.~\eqref{eq:transport-triangular}.
  \item \emph{Strict monotone coordinates:} $\partial_{u_k} S_k(u_{1:k}) > 0$ for all $k$ and all arguments.
  \item \emph{Separable component:} Eq.~\eqref{eq:transport-separable} holds, so conditional shape along $u_k$ is fixed across contexts.
\end{itemize}
\end{shaded}

>>>>

Multivariate density estimation is a cornerstone of modern statistics and machine learning. It supplies the machinery for likelihood-based inference, simulation of generative mechanisms, and uncertainty quantification in tabular domains. Practitioners, however, routinely face a delicate trade-off between three goals: expressive power, computational feasibility, and interpretability. Three complementary families of estimators have emerged around this dilemma:
\begin{enumerate}
  \item \textbf{Normalizing flows} build expressive change-of-variables maps and have become central in deep generative modeling \citep{rezende2015variational,papamakarios2017masked,dinh2017real,durkan2019neural,kingma2018glow,papamakarios2021normalizing}.
  \item \textbf{Transformation models and forests} model conditional distributions through parametric transformation structures coupled with ensemble methods \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation}.
  \item \textbf{Copulas} separate marginals from dependence, yielding interpretable multivariate association models \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence,nagler2017kdecopula}.
\end{enumerate}

This thesis situates all three families inside a common \textbf{transport framework} that makes them directly comparable \citep{rosenblatt1952remarks,knothe1957contributions,bogachev2005triangular,ramgraber2025friendly}. The framework operates in standardized coordinates $u$ computed solely from training-split statistics to avoid leakage and keep objectives aligned across models. Within this evaluation framework each estimator couples $u$ to a reference distribution---Gaussian for triangular maps and TRTF, $\mathrm{Unif}([0,1]^K)$ for copulas---and recovers densities via exact change of variables. Section~\ref{sec:ch1-frame} previews the core pieces and Chapter~\ref{ch:background} develops the mathematics in detail.

Separable triangular transport maps play a central role throughout the thesis. Their lower-triangular Jacobian delivers a log-determinant in $\mathcal{O}(K)$ operations, back-substitution yields an explicit inverse for sampling, and the triangular structure produces exact conditional densities in a fixed order. Fully triangular maps can require $\mathcal{O}(K^2)$ work per evaluation, whereas separable variants reduce both forward and inverse passes to $\mathcal{O}(K)$, making them attractive for systematic comparison while preserving interpretable marginals and dependence mechanisms.

\medskip
\noindent\textit{Notation.} We write $\varphi$ and $\Phi$ for the density and cumulative distribution function of a univariate standard normal random variable, and $\eta$ for the $K$-variate standard normal density. All models are evaluated in standardized coordinates $u$ and reported on the original scale $x$ through a diagonal affine correction so that log densities remain comparable. Unless stated otherwise, log values are expressed in natural units (nats).

\section{Thesis and Problem Statement}\label{sec:ch1-problem}

The goal of this thesis is to examine multivariate density estimation for tabular data within a unified transport-based evaluation protocol. We focus on three estimators: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and nonparametric copula models.

Let $u = D^{-1}(x - \mu)$ with $D = \mathrm{diag}(\sigma_1,\ldots,\sigma_K)$ determined from the training split. If $T: \mathbb{R}^K \to \mathbb{R}^K$ is lower triangular, $z = T(u)$, and $\eta$ denotes the chosen reference density, then the change-of-variables formula gives
\begin{equation}
  \log p_X(x) = \log \eta\big(T(u)\big) + \sum_{k=1}^{K} \log \frac{\partial T_k}{\partial u_k}(u_{1:k}) - \sum_{k=1}^{K} \log \sigma_k,
  \label{eq:ch1-change-of-variables}
\end{equation}
so reporting on the $x$-scale differs from reporting on the $u$-scale by the constant $-\sum_k \log \sigma_k$ per data point. The Rosenblatt and Knothe rearrangements guarantee the existence of such triangular transports for any coordinate ordering \citep{rosenblatt1952remarks,knothe1957contributions}.

The copula baselines considered here are nonparametric kernel copulas \citep{nagler2017kdecopula}. We fit smooth univariate marginals with \texttt{kde1d}, form pseudo-observations using mid-rank empirical CDFs during training to avoid double smoothing, and evaluate joint densities through
\[
  p_X(x) = c\big(F_1(x_1),\ldots,F_K(x_K)\big)\, \prod_{k=1}^{K} f_k(x_k),
\]
where $c$ is the copula density and $F_k = \int f_k$ the smoothed marginal CDFs. No Gaussian copula enters the study; the emphasis remains on flexible, interpretable diagnostics in low dimensions.

Separable triangular maps decompose each coordinate into a context-dependent shift plus a one-dimensional monotone transform. TRTF occupies an intermediate position: under strictly increasing conditional CDFs and the standard forest aggregation after monotone smoothing it reproduces the same separable triangular likelihood via the probability integral transform. The shared protocol described above keeps preprocessing, likelihood evaluation, calibration diagnostics, and compute measurements aligned across all estimators.

\section{The Transport Frame on One Page}\label{sec:ch1-frame}

The key ingredients of the transport framework are:
\begin{itemize}
  \item \textbf{Standardization.} Coordinates $u = D^{-1}(x-\mu)$ use training-split statistics only; all logs are in nats.
  \item \textbf{Change of variables.} Densities follow from $z = T(u)$ through Equation~\eqref{eq:ch1-change-of-variables}.
  \item \textbf{Triangular log determinant.} Lower-triangular Jacobians imply $\log \det J_T(u) = \sum_k \log \partial_{u_k} T_k(u_{1:k})$, giving $\mathcal{O}(K)$ cost.
  \item \textbf{Affine correction.} Reporting on $x$ applies the constant correction $-\sum_k \log \sigma_k$ per data point.
  \item \textbf{Computational profile.} Fully triangular maps can incur $\mathcal{O}(K^2)$ evaluation and inversion, whereas separable maps reduce both passes to $\mathcal{O}(K)$ while keeping the log determinant linear in $K$.
\end{itemize}

Separable triangular transports implement each coordinate as the sum of a context shift and a monotone one-dimensional basis, maintaining determinant stability and consistent conditional behaviour. TRTF achieves the same likelihood under the conditions noted above, while nonparametric copulas operate on the unit hypercube with explicit marginals. Standardization $u = T_{\mathrm{std}}(x)$ feeds either the triangular transport branch (TTM variants, TRTF) or the copula branch, and both return log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.

\section{Contributions and Research Questions}\label{sec:ch1-contributions}

The thesis contributes along three axes that move from theoretical unification to empirical benchmarking and practical interpretation:
\begin{enumerate}
  \item \textbf{Unified likelihood framework.} We formalise a shared transport perspective that places separable triangular maps, TRTF, and nonparametric copulas inside one mathematical scheme. The equivalence between TRTF and separable triangular transports holds under explicit regularity assumptions---strictly monotone conditional CDFs and the forest aggregation after smoothing.

    \begin{quote}
    \textbf{Proposition (informal).} Under axis-parallel partitions, strictly monotone conditional CDFs, and the forest aggregation after smoothing, the TRTF likelihood coincides with that of a separable triangular transport in standardized coordinates.
    \end{quote}

  \item \textbf{Benchmarking under a common protocol.} We design a standardized evaluation pipeline covering synthetic generators and real tabular benchmarks. TTM-Sep, its marginal variant, and TRTF are evaluated systematically; nonparametric kernel copulas act as diagnostic comparators when $K \le 3$. The pipeline produces three categories of evidence: (i) average test log likelihoods, (ii) conditional diagnostics based on probability integral transforms, and (iii) compute measurements for training effort and per-sample evaluation time. Together they quantify the trade-offs among accuracy, calibration, and efficiency.

  \item \textbf{Operational insights.} The theoretical and empirical findings yield guidance for practitioners. We analyse sensitivity to ordering, the practical limits of separability, and when nonparametric copulas remain informative as diagnostic references. Chapter~\ref{ch:conclusion} synthesizes these insights and outlines extensions toward richer parameterisations.
\end{enumerate}

Two research questions organise the empirical study:
\begin{itemize}
  \item \textbf{Synthetic data.} How do TRTF, TTM-Sep, and nonparametric copulas compare on controlled generators? Evidence comprises likelihood comparisons, conditional calibration metrics, and timing measurements.
  \item \textbf{Benchmark datasets.} How close do TRTF results come to published normalizing flow baselines under identical preprocessing? Side-by-side likelihood comparisons, interpreted through the constraints of separability and compute profiles, answer this question.
\end{itemize}

\begin{table}[t]
  \centering
  \caption{Model abbreviations used throughout the thesis.}
  \label{tab:model-abbrev}
  \begin{tabular}{ll}
    \hline
    Label & Meaning \\
    \hline
    TTM-Marg  & Marginal triangular transport (per-dimension; no context) \\
    TTM-Sep   & Separable triangular transport (context shift $+$ monotone 1D transform) \\
    TRTF      & Transformation Random Forests (axis-parallel splits) \\
    True-Marg & Oracle marginal density \\
    True-Joint& Oracle conditional joint density \\
    Copula    & Nonparametric kernel copula baseline (smoothed marginals; no Gaussian copula) \\
    \hline
  \end{tabular}
\end{table}

Addressing these questions bridges rigorous transport-based theory with empirical validation, clarifying when simple triangular models suffice and when more expressive families become necessary.

@@

\section{Transport Framework and Notation}\label{sec:transport-frame}

This section establishes the coordinate system, notation, and algebraic identities used throughout the thesis. The schematic in Appendix~\ref{ch:appendix} remains valid; here we focus only on the formulas that are needed later. We set up the standardized pullback likelihood, the triangularity assumption, and the Jacobian factorization.

We write $u_{1:k}=(u_1,\ldots,u_k)$ and use $\partial_{u_k}$ for partial derivatives with respect to $u_k$.

We observe data on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
\begin{equation}
  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,
  \label{eq:transport-standardise}
\end{equation}
where $\mu$ and $\sigma$ are the empirical mean and standard deviation estimated on the training split, and $\oslash$ denotes elementwise division. Each feature is shifted and rescaled once using training data only. All derivatives and Jacobians are then taken in $u$-space, which avoids leakage and keeps results comparable across estimators.

We model $\pi_U$ as the pullback of a simple reference $\eta$ through a triangular map $S:u\mapsto z$. The reference is the $K$-variate standard normal $\eta(z)$. Applying the change-of-variables formula gives
\begin{equation}
  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|.
  \label{eq:transport-pullback}
\end{equation}

To report log densities on the original scale, we apply only the diagonal affine correction implied by standardization:
\begin{equation}
  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.
  \label{eq:transport-affine}
\end{equation}

The transport is assumed to be lower triangular and strictly increasing in each coordinate,
\begin{equation}
  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), 
  \qquad \partial_{u_k}S_k(u_{1:k})>0,
  \label{eq:transport-triangular}
\end{equation}
so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorizes as
\begin{equation}
  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).
  \label{eq:transport-det}
\end{equation}

This factorization makes the \emph{log-determinant} cost $\mathcal{O}(K)$ once the diagonal terms are available. Evaluating all $S_k(u_{1:k})$ can be $\mathcal{O}(K^2)$ for a general triangular map, but reduces to $\mathcal{O}(K)$ in the separable parameterization of Section~\ref{sec:transport-separable}.

Strict monotonicity on each coordinate ($\partial_{u_k} S_k>0$) together with triangularity implies that $S$ is a bijection with a triangular inverse, computable by back-substitution.

\begin{table}[t]
  \centering
  \caption{Notation for the transport framework used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}. All derivatives and Jacobians are with respect to $u$; log densities on $x$ apply the affine correction in Equation~\eqref{eq:transport-affine}.}
  \label{tab:transport-notation}
  \begin{tabular}{ll}
    \hline
    Symbol & Meaning \\ 
    \hline
    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ \\ 
    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates \\ 
    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
    $\eta(z)$ & $K$-variate standard normal density \\ 
    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space \\ 
    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
    $K$ & Dimension of the feature vector \\ 
    \hline
  \end{tabular}
\end{table}

Taken together, the triangular and strictly monotone structure turns joint density evaluation
and sampling into sequential one-dimensional stages, and the standardized coordinates allow us
to report on the $x$-scale with only the diagonal correction from Equation~\eqref{eq:transport-affine}.

\section{Separable Triangular Maps and Transformation Random Forests}\label{sec:transport-separable}

We now move from the general triangular setup to the first concrete parameterization: the \textbf{separable triangular map}. We then show how \textbf{Transformation Random Forests (TRTF)} fit into the same framework. The aim is a shared likelihood that we can train, compare, and diagnose across estimators without changing conventions.

\subsection{Separable triangular component}

A separable component has the form
\begin{equation}
  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k), \qquad h_k'(u_k) > 0,
  \label{eq:transport-separable}
\end{equation}
where earlier coordinates shift location through $g_k$, while $h_k$ shapes the marginal and ensures monotonicity. To fix identifiability, we enforce a convention such as centering $g_k$ on the training split, $\mathbb E[g_k(U_{1:k-1})]=0$.

With the standard normal reference, the pullback identity becomes
\begin{equation}
  \log \pi_U(u) \;=\; \sum_{k=1}^{K}\Big[\log \varphi\!\big(S_k(u_{1:k})\big) + \log h_k'(u_k)\Big].
  \label{eq:transport-likelihood}
\end{equation}
Up to a constant, the per-sample negative log-likelihood is
\begin{equation}
  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big].
  \label{eq:transport-loss}
\end{equation}

For a dataset $\{u^{(i)}\}_{i=1}^n$, the empirical objective reads
\[
\mathcal L_n = \frac{1}{n}\sum_{i=1}^n \sum_{k=1}^{K}
\Big[\tfrac{1}{2}\,S_k(u^{(i)}_{1:k})^2 - \log h_k'(u^{(i)}_k)\Big] + \text{const}.
\]

These formulas evaluate in $\mathcal{O}(K)$ time per sample, since the Jacobian term depends only on $u_k$. Sampling inverts $S$ by back-substitution, and log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}. The speed comes with a trade-off: separability fixes the conditional shape along $u_k$ once $g_k$ shifts the location, so variance, skewness, and modality no longer adapt to the context $u_{1:k-1}$.

\subsection{Transformation Random Forests in the framework}\label{sec:transport-trtf}

Transformation Random Forests (TRTF) fit conditional CDFs and insert them into the triangular pipeline via the probability integral transform. Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the smoothed conditional CDF. Define
\begin{equation}
  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big).
  \label{eq:transport-trtf-map}
\end{equation}

If the forest is parameterized additively as
\begin{equation}
  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),
  \label{eq:transport-trtf-additive}
\end{equation}
then we recover the separable structure,
\begin{equation}
  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}), \qquad h_k'(u_k) > 0.
  \label{eq:transport-trtf-separable}
\end{equation}

\paragraph{Derivative check.}
\[
\partial_{u_k}S_k
= \frac{d}{du_k}\Phi^{-1}(\Phi(h_k(u_k)+g_k))
= h_k'(u_k) > 0,
\]
so the Jacobian diagonal is indeed $h_k'(u_k)$ in the additive TRTF case.

\paragraph{Numerical note.}
In practice, forests may produce $\widehat F_k$ values outside $(0,1)$ by tiny amounts or with imperfect monotonicity. We therefore clip to $(\varepsilon,1-\varepsilon)$ with $\varepsilon=10^{-6}$ and, if needed, apply isotonic post-processing to enforce strict increase. This keeps $S_k=\Phi^{-1}(\widehat F_k)$ well defined.
